---
title: "Dominance Analysis"
author: "Adon Rosen"
date: 'Date: `r Sys.Date()`'
output: pdf_document
---
  
```{r, echo=F, message=F, warning=FALSE}
library(knitr)
opts_knit$set(root.dir='./')  #Don't combine this call with any other chunk -especially one that uses file paths.
```


<!-- Set the report-wide options, and point to the external code file. -->
```{r set_options, echo=F, warning=FALSE}
# cat("Working directory: ", getwd())
report_render_start_time <- Sys.time()
opts_chunk$set(
  results    = 'show',
  comment    = NA,
  tidy       = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.path   = 'figure-png/'
)
# echoChunks <- FALSE
options(width=80) #So the output is 50% wider than the default.
read_chunk("./dominanceAnalysis.R") #This allows knitr to call chunks tagged in the underlying *.R file.
```

```{r load-packages, echo=FALSE, results='show', message=FALSE, warning=FALSE}
```


# Question 1 - Purpose of a Dominance Analysis:

  The dominance analysis, as described by Azen and Budescu (2003), was proposed to be more intuitive and limit possible misinterpretations when exploring variable importance in multiple regression models. The authors argue this by first presenting commonly used techniques used to analyze variable importance including standardized coefficients, squared partial correlations, and commonality analysis as well as the limitations of these techniques such as susceptibility to the influence of multicollinearity, order dependent effects, and difficulty interpreting when higher order interactions are present. In order to overcome these limitations the authors reintroduce Dominance Analysis (DA). DA is a technique proposed by Budescu (1993) which allows for the relative importance of a variable to be explored through all $2^{(p-2)}$ possible model permutations. This exhaustive procedure allows for the growth of a model to be explored throughout the addition of individual variables. This procedure affords insight into both the predictive and inferential processes of multiple regression. It can be used for the predictive application by finding possible variance explained across all models and attempting to find subsets that satisfy the experimenters desires. This allows for more affordable bias-variance tradeoffs by exploring for possible plateaus in additional variance explained throughout the model building procedure. It benefits inferential approaches by informing an experimenter when a variable can describe portions of variance and also when multicollinearity may be an issue among variables.

  The authors also introduced language in order to benefit potential inferential procedures such as concepts of complete dominance, relative importance, conditional dominance, and also general dominance. Complete dominance details variables that across all model folds are more dominant than another variable. If there were three possible variables $X_1$ would be completely dominant over $X_3$ if the variance explained in models consisting of $X_1$ and $X_1 + X_2$ explain more variance than the same permutations exchanging $X_1$ for $X_3$. Relative importance details the additional variance explained when a variable is added to a model. Conditional dominance explains how a variable has a greater relative importance when the model is conditioned on other variables. Finally, general dominance details how all possible permutations of a model at a specific level explain greater variance than other possible variable permutations.







# Question 2 - Example and explanation of a Dominance Analysis:

```{r print-da, echo=FALSE, results='show', message=FALSE, warning=FALSE}
```

This table details the variance explained when models were fitted predicting crime rate as functions of: average income for an area; number of individuals with a high school diploma; and finally how urban an area was (population density). Here we see that Urbanicity is completely dominant, High school is conditionally dominant over income at level 2, and is generally dominant over income as well at level 1 and 2.



# Bibliography

Azen, R., & Budescu, D. V. (2003). The dominance analysis approach for comparing predictors in multiple regression. Psychological Methods, 8(2), 129–148. https://doi.org/10.1037/1082-989x.8.2.129


Budescu, D. V. (1993). Dominance analysis: A new approach to the problem of relative importance of predictors in multiple regression. Psychological Bulletin, 114(3), 542–551. https://doi.org/10.1037/0033-2909.114.3.542